---
title: "Assignment3 - Introuction to Data Analysis"
author: "Sheetal Gangakhedkar"
date: "May 8, 2017"
output:
  word_document:
    toc: yes
    toc_depth: 4
  pdf_document: default
  html_document:
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Import the data into R.

###1a) How many observations are there in the dataset?
```{r}
# load the forest fires data from the csv file at
# https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/forestfires.csv
# getwd()
# setwd("C:/Users/esheega/Documents/UCSC/DataAnalyticsR/assignment3")
ffires = read.csv("forestfires.csv")
dim(ffires)[1]
```

###1b) How many observations are there with a fire (i.e., area>0)
```{r}
areaFires = ffires[which(ffires$area > 0.0),]
dim(areaFires)[1]
```

###1c) How many observations are there with a rain (i.e., rain>0)
```{r}
rainFires = ffires[which(ffires$rain > 0.0),]
dim(rainFires)[1]
```

###1d) How many observations are there with both a fire and a rain?
```{r}
rainAreaFires = ffires[which(ffires$rain > 0.0 & ffires$area > 0.0),]
dim(rainAreaFires)[1]
```

##2) Show the columns month, day, area of the all the observations.
```{r}
ffires[,c("month","day","area")]
```

##3) Show the columns month, day, area of the observations with a fire.
```{r}
ffires[which(ffires$area > 0.0),c("month","day","area")]
```

##4) How large are the five largest fires (i.e., having largest area).
```{r}
head(ffires[with(ffires, order(-area)),], 5)$area
```
###4a) What are the corresponding month, temp, RH, wind, rain area?
```{r}
head(ffires[with(ffires, order(-area)),], 5)[,c("month","temp","RH", "wind", "rain", "area")]
```
###4b) Reorder factor levels of month to be from Jan to Dec.
```{r}
levels(ffires$month)
ffires$month = factor(ffires$month, levels=c("jan","feb","mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))
```
```{r}
levels(ffires$month)
```
###4b) Add one column to the data indicating whether a fire occured for each observation (true for area > 0 and false for area == 0)
```{r}
ffires$fire = if (ffires$area > 0.0) TRUE else FALSE
head(ffires)
```